Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-04-27 LICS2021-final-with-appendices-for-arxiv.tex
improved description of Fig 1 examples, per referee request

This is where \emph{combinatorial proofs} come in. They were
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a syntax-free notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in inductively-generated proofs
and consider the proof as a combinatorial object, more precisely
as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula=\peirceformula$ is shown mid-left in
Fig.\,\ref{fig:translate},
%depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
\emph{drinker paradox} \mbox{$\drinkerformula$} $=$ \mbox{$\veedrinkerformula$}
is shown on the right of Fig.\,\ref{fig:translate}, 
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is unsatisfactory: (1) the soundness argument is long, intricate
and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} nor
\emph{full}~\cite{abramsky:jagadeesan:94}. 
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

>>>>

This is where \emph{combinatorial proofs} come in. They were
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a syntax-free notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in inductively-generated proofs
and consider the proof as a combinatorial object, more precisely
as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula=\peirceformula$ is shown mid-left in
Fig.\,\ref{fig:translate},
%depicted below-left,
a homomorphism from a 4-vertex graph with two colours (above) to a 4-vertex graph labelled with
propositional variables (below); dotted vertical lines define the homomorphism, from above to below.
%

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
\emph{drinker paradox} \mbox{$\drinkerformula$} $=$ \mbox{$\veedrinkerformula$}
is shown on the right of Fig.\,\ref{fig:translate}, 
a homomorphism from a 5-vertex partially coloured graph to a 4-vertex labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is unsatisfactory: (1) the soundness argument is long, intricate
and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} nor
\emph{full}~\cite{abramsky:jagadeesan:94}. 
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.
